Aneuploidy may be an evolutionary sidetrack on the path to adaptation

### 2022-07-26 ms/ms.tex
added some todos
ack DBW

@@

% Fig dynamics
\begin{figure}[h!]
      \centering
      \includegraphics[width=\textwidth]{../figures/FA.pdf}      
  \caption{
    \textbf{Predicted frequency of aneuploid-descended cells.}     
    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$. Colored curves are 50 simulations using the MAP estimate parameters. Black dashed curves are the expected genotype frequencies without genetic drift. 
     See \Cref{fig:MAP_freq_dynamics_log} for log-log scale, in which the sequence of events is easier to observe.
    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
    Solid and dashed lines show the mean and 95\% CI.
	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
	\textbf{(D)} $F_A$ as a function of the population size, $N$, in posterior predictions with MAP parameters. Markers show $F_A$ in 250 simulations per population size. Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$). Vertical dashed line for population size in the experiment, $6.425\cdot10^6$. Horizontal line for $F_A^{MAP}=0.106$. 
     }
  \label{fig:FA}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

This happens despite aneuploidy reaching a high frequency in the population (>95\%). 
Conventional wisdom might suggest that once the aneuploid genotype \anwt\ reaches high frequency, it will have a better chance at producing "refined" solutions via mutations, and its descendants will come to dominate the population: the frequency \eumtA\ (which arises from \anmt) will be higher than the frequency of \eumtM\ (which arises directly from \euwt).
So how does \eumtM\ prevail?
At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
Accordingly, \anmt\ cells appear more frequently than \eumtM. However, the fitness advantage of a "refined" mutation that emerges in the \anmt\ genotype in hindered by the cost of aneuploidy as well as the extra waiting time needed to eliminate the extra copy of the chromosome by a chromosome loss event.
Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
Indeed, we estimate that the advantage of \anmt\ over \anwt\ is similar to the advantage of \eumt\ over \anmt (about 0.3\%), and that the rate of chromosome loss is roughly $1-3\cdot10^{-3}$.

@@

% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, and Tal Simon for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (ISF, YR 552/19),
the US--Israel Binational Science Foundation (BSF, YR 2021276)
Minerva Stiftung center for Lab Evolution (YR),
Minerva Stiftung short-term research grant (MP).
}

>>>>

% TODO F_A vs N

% TODO log-log scale of freq dynamics

@@

% Fig dynamics
\begin{figure}[h!]
      \centering
      \includegraphics[width=0.9\textwidth]{../figures/FA.pdf}      
  \caption{
    \textbf{Predicted frequency of aneuploid-descended cells.}     
    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$. Colored curves are 50 simulations using the MAP estimate parameters. Black dashed curves are the expected genotype frequencies without genetic drift. 
     See \Cref{fig:MAP_freq_dynamics_log} for log-log scale, in which the sequence of events is easier to observe.
    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
    Solid and dashed lines show the mean and 95\% CI.
	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
	\textbf{(D)} $F_A$ as a function of the population size, $N$, in posterior predictions with MAP parameters. Markers show $F_A$ in 250 simulations per population size. Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$). Vertical dashed line for population size in the experiment, $6.425\cdot10^6$. Horizontal line for $F_A^{MAP}=0.106$. 
     }
  \label{fig:FA}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

This happens despite aneuploidy reaching a high frequency in the population (>95\%). 
Conventional wisdom might suggest that once the aneuploid genotype \anwt\ reaches high frequency, it will have a better chance at producing "refined" solutions via mutations, and its descendants will come to dominate the population: the frequency \eumtA\ (which arises from \anmt) will be higher than the frequency of \eumtM\ (which arises directly from \euwt).
So how does \eumtM\ prevail?
% TODO log-log scale of freq dynamics
At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
Accordingly, \anmt\ cells appear more frequently than \eumtM. However, the fitness advantage of a "refined" mutation that emerges in the \anmt\ genotype in hindered by the cost of aneuploidy as well as the extra waiting time needed to eliminate the extra copy of the chromosome by a chromosome loss event.
Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
Indeed, we estimate that the advantage of \anmt\ over \anwt\ is similar to the advantage of \eumt\ over \anmt (about 0.3\%), and that the rate of chromosome loss is roughly $1-3\cdot10^{-3}$.

@@

% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, Daniel Weissman, and Tal Simon for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (ISF, YR 552/19),
the US--Israel Binational Science Foundation (BSF, YR 2021276)
Minerva Stiftung center for Lab Evolution (YR),
Minerva Stiftung short-term research grant (MP).
}
